---
title: "Getting started with fmriphysio"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with fmriphysio}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(fmriphysio)
pct <- function(x) sprintf("%.0f%%", 100 * x)
```

## The problem

Cardiac pulsation and respiration add structured variance to BOLD time
series. The signals are strongest in and near large vessels and
cerebrospinal fluid, but they reach grey matter too. At conventional
repetition times they are also aliased: a heart rate near 1 Hz sampled every
2 s turns up as a slow oscillation that a temporal filter can't separate from
neural signal. Methods that model the noise from external recordings, such as
RETROICOR, need a pulse oximeter and a respiratory belt, and those recordings
are often missing or unusable.

`fast_phy_denoise()` estimates the noise from the fMRI data alone and needs no
task design. It follows the logic of PHYCAA+ (Churchill & Strother, 2013):

1. Give each voxel a neuronal-tissue weight `wNN` (1 = neuronal,
   0 = non-neuronal) from the energy of its temporal differences.
2. Weight the data toward non-neuronal voxels, reduce them to a small number
   of temporal components, and extract candidate components by lag-1
   canonical autocorrelation analysis, which orders them by how well they
   predict their own next time point.
3. Keep a candidate as noise when its median voxelwise R² in non-neuronal
   voxels is more than twice that in neuronal voxels.
4. Project the kept components out of every voxel in one step, restoring voxel
   means.

Use it for resting-state or task data that have no usable physiological
recordings, before connectivity or GLM analysis. Don't use it on data that
have already been physiologically corrected (see
[Limitations](#limitations)).

## Simulated data with a known answer

To check a denoiser you need to know which variance is physiological.
`simulate_phy_data()` returns a voxels × time matrix together with its
sources: two physiological sources (cardiac and respiratory, defined in Hz
and sampled at the TR, so they alias as real signals do), three slow neural
sources, the voxel loadings, and the true tissue label of each voxel. Voxel
means sit near 100, so temporal SNR is on a realistic scale.

```{r simulate}
sim <- simulate_phy_data(n_vox = 1500, n_time = 300, tr = 2, seed = 1)
sim
```

At this TR the Nyquist frequency is 0.25 Hz. The frequencies the base
physiological rates alias to are:

```{r alias}
alias_hz <- abs(sim$physio_hz - round(sim$physio_hz * sim$tr) / sim$tr)
round(alias_hz, 3)
```

The cardiac source lands near `r round(alias_hz[["cardiac"]], 3)` Hz, below
the 0.08 Hz cutoff of the simulated neural sources. The simulated rates drift
slowly, so the aliased power spreads around these frequencies, but either way
a band-pass filter can't remove the cardiac noise without removing neural
signal too.

## Denoise

`fast_phy_denoise()` needs only the data. The `tr` argument is stored as
metadata; the algorithm doesn't use it.

```{r fit}
fit <- fast_phy_denoise(sim$x, tr = sim$tr)
fit
```

`fit$x_clean` has the same shape as `sim$x`. `fit$regressors` holds the time
courses that were projected out, and `fit$component_table` explains why each
candidate was kept or not:

```{r component-table}
ct <- fit$component_table
ct$ratio_nn_nt <- round(ct$ratio_nn_nt, 2)
ct$predictability <- round(ct$predictability, 3)
head(ct[, c("candidate", "ratio_nn_nt", "predictability", "selected")], 8)
```

Candidates are ordered by predictability, the canonical lag-1 correlation.
The two selected candidates are expressed more than ten times as strongly in
non-neuronal as in neuronal voxels (`ratio_nn_nt`). Candidates 3 to 5 are also
highly predictable, but their ratios are far below 1. They are expressed
mainly in neuronal tissue, so they're kept. Predictability alone doesn't
separate noise from signal. The spatial ratio does.

Here the ground truth confirms it. Each removed regressor matches one
physiological source:

```{r regressors-vs-truth}
round(abs(cor(fit$regressors, sim$physio)), 2)
```

## Is the tissue map right?

The whole method rests on `wNN`, so compare it with the true tissue labels:

```{r wnn-check}
tissue <- factor(ifelse(sim$nn, "non-neuronal", "neuronal"))
table(tissue, wNN = ifelse(fit$wNN < 0.5, "< 0.5", "> 0.5"))
round(tapply(fit$wNN, tissue, mean), 3)
```

The default map is conservative. No neuronal voxel is labelled non-neuronal,
but `r sum(sim$nn & fit$wNN > 0.5)` of the `r sum(sim$nn)` non-neuronal
voxels get `wNN > 0.5`. The weight ramps from 1 at the 74th percentile of
excess difference energy (`delta_nt = 0.74`) to 0 at the 95th
(`delta_nn = 0.95`), so only the upper tail is treated as non-neuronal.
Because selection compares medians, a partly labelled non-neuronal set is
enough here.

## How much noise went, and how much signal stayed?

With the sources known, we can measure, in each truly neuronal voxel, the
fraction of the energy in the span of the physiological sources (or of the
neural sources) that survives denoising: 0 means all removed, 1 means all
kept. This helper is only for evaluation; it isn't part of the package.

```{r energy-helper}
energy_kept <- function(x_raw, x_clean, sources) {
  Q <- qr.Q(qr(scale(sources, scale = FALSE)))
  before <- rowSums(((x_raw - rowMeans(x_raw)) %*% Q)^2)
  after <- rowSums(((x_clean - rowMeans(x_clean)) %*% Q)^2)
  after / before
}
kept <- function(sim, x_clean) {
  nt <- !sim$nn
  c(physio = median(energy_kept(sim$x[nt, ], x_clean[nt, ], sim$physio)),
    neural = median(energy_kept(sim$x[nt, ], x_clean[nt, ], sim$neural)))
}
```

CompCor (Behzadi et al., 2007) is the usual data-driven alternative:
it removes the leading principal components of a set of nuisance voxels.
`compcor_denoise()` implements it as a baseline. `"acompcor"` takes its
nuisance voxels from `nuisance_mask` or, as here, from the voxels with
`wNN < 0.5`. `"tcompcor"` takes the highest-variance voxels. Both remove a
fixed number of components, `n_comp` (default 5).

```{r compare}
cc_a5 <- compcor_denoise(sim$x, tr = sim$tr, mode = "acompcor")
cc_a2 <- compcor_denoise(sim$x, tr = sim$tr, mode = "acompcor", n_comp = 2)
cc_t5 <- compcor_denoise(sim$x, tr = sim$tr, mode = "tcompcor")
round(rbind(
  fast_phy_denoise     = kept(sim, fit$x_clean),
  "aCompCor, n_comp 5" = kept(sim, cc_a5$x_clean),
  "aCompCor, n_comp 2" = kept(sim, cc_a2$x_clean),
  "tCompCor, n_comp 5" = kept(sim, cc_t5$x_clean)
), 3)
```

`fast_phy_denoise()` removes almost all of the physiological energy from
neuronal voxels and keeps almost all of the neural energy. All three CompCor
variants remove the physiological energy too, but with five components aCompCor
also removes nearly all of the neural energy, and tCompCor removes most of it.
In this simulation the non-neuronal voxels carry weak neural signal. The
comparison of `n_comp = 2` with `n_comp = 5` shows that the three components
beyond the second carry it, and fixed-rank CompCor removes them from every
voxel. With `n_comp = 2`, the true number of physiological sources, aCompCor
does as well as `fast_phy_denoise()`. So CompCor isn't inherently worse here,
but it depends on a rank chosen in advance, whereas `fast_phy_denoise()`
selects components by where they are expressed.

The spectra show where the removed variance was:

```{r spectra, fig.width = 7, fig.height = 3.4, fig.alt = "Mean power spectra of non-neuronal and neuronal voxels before and after denoising. In non-neuronal voxels the physiological power near both aliased frequencies is removed; in neuronal voxels the respiratory peak is removed and the spectrum below 0.1 Hz, dominated by neural signal, looks unchanged."}
mean_spectrum <- function(x, tr) {
  z <- t(x - rowMeans(x))
  k <- seq_len(floor(nrow(z) / 2))
  p <- rowMeans(Mod(mvfft(z))^2) / nrow(z)
  list(hz = k / (nrow(z) * tr), power = p[k + 1])
}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1), cex = 0.8)
for (lab in c("non-neuronal", "neuronal")) {
  v <- tissue == lab
  before <- mean_spectrum(sim$x[v, ], sim$tr)
  after <- mean_spectrum(fit$x_clean[v, ], sim$tr)
  plot(before$hz, before$power, type = "l", log = "y", col = "grey55",
       xlab = "Frequency (Hz)", ylab = "Mean power", main = paste(lab, "voxels"))
  lines(after$hz, after$power, col = "firebrick")
  abline(v = alias_hz, lty = 3)
  if (lab == "non-neuronal") {
    legend("top", c("raw", "denoised"), col = c("grey55", "firebrick"),
           lty = 1, bty = "n")
  }
}
par(op)
```

Dotted lines mark the frequencies the base cardiac and respiratory rates alias
to. In non-neuronal voxels the physiological power around both lines is gone.
In neuronal voxels the respiratory peak is removed. Below about 0.1 Hz their
spectrum is dominated by neural signal and looks unchanged: the cardiac noise
aliased into that band is too weak there to see, although the table above
shows that the physiological energy is removed.

## Across datasets and TRs

One dataset is an anecdote. The next chunk repeats the comparison over six
simulated datasets at each of two TRs. It also runs the settings closer to
the PHYCAA+ specification (weighting toward neuronal tissue, a rank-100
subspace, and a ratio threshold of 1), which the package defaults depart
from.

```{r replicate}
phycaa_like <- list(extraction_weight = "neuronal", pca_rank = 100,
                    ratio_thresh = 1)
runs <- do.call(rbind, lapply(c(0.8, 2), function(tr) {
  do.call(rbind, lapply(1:6, function(seed) {
    s <- simulate_phy_data(n_vox = 1500, n_time = 300, tr = tr, seed = seed)
    rbind(
      data.frame(tr = tr, seed = seed, settings = "defaults",
                 t(kept(s, fast_phy_denoise(s$x)$x_clean))),
      data.frame(tr = tr, seed = seed, settings = "PHYCAA+-like",
                 t(kept(s, do.call(fast_phy_denoise,
                                   c(list(s$x), phycaa_like))$x_clean)))
    )
  }))
}))
summary_tab <- do.call(rbind, lapply(split(runs, runs[c("settings", "tr")]),
  function(d) data.frame(
    settings = d$settings[1], tr = d$tr[1],
    physio_mean = mean(d$physio), physio_worst = max(d$physio),
    neural_mean = mean(d$neural), neural_worst = min(d$neural)
  )))
rownames(summary_tab) <- NULL
format(summary_tab, digits = 2)
```

```{r replicate-summary, include = FALSE}
def <- summary_tab[summary_tab$settings == "defaults", ]
```

With the defaults, on average about `r pct(def$physio_mean[def$tr == 0.8])`
(TR 0.8 s) and `r pct(def$physio_mean[def$tr == 2])` (TR 2 s) of the
physiological energy remains in neuronal voxels, and at least
`r pct(min(def$neural_worst))` of the neural energy is kept in every run. The
PHYCAA+-like settings leave more physiological energy and remove more neural
energy. The average hides a weaker run at TR 2 s:

```{r worst-run}
runs[runs$settings == "defaults" & runs$tr == 2, c("seed", "physio", "neural")]
```

Break the worst run (seed 2) down by source:

```{r worst-run-sources}
s2 <- simulate_phy_data(n_vox = 1500, n_time = 300, tr = 2, seed = 2)
fit2 <- fast_phy_denoise(s2$x)
nt2 <- !s2$nn
by_source <- rbind(
  alias_hz = abs(s2$physio_hz - round(s2$physio_hz * s2$tr) / s2$tr),
  energy_kept = sapply(colnames(s2$physio), function(j) {
    median(energy_kept(s2$x[nt2, ], fit2$x_clean[nt2, ],
                       s2$physio[, j, drop = FALSE]))
  }),
  best_abs_r = apply(abs(cor(fit2$regressors, s2$physio)), 2, max)
)
round(by_source, 2)
```

Here `r ncol(fit2$regressors)` components are removed, but
`r pct(by_source["energy_kept", "cardiac"])` of the cardiac energy remains in
neuronal voxels, and no removed regressor matches the cardiac source closely
(best |r| = `r round(by_source["best_abs_r", "cardiac"], 2)`). In this run the
two sources alias to nearby frequencies
(`r round(by_source["alias_hz", "cardiac"], 2)` and
`r round(by_source["alias_hz", "respiratory"], 2)` Hz), and the candidates
that pass selection capture the cardiac source only in part. Inspect
`component_table` and `regressors` whenever the result matters.

## Design-free quality control

On real data there's no ground truth. `compute_design_free_qc()` summarises
what denoising changed, separately for voxels with `wNN > 0.5` (neuronal) and
`wNN < 0.5` (non-neuronal). Temporal SNR after denoising is computed with the
raw voxel mean, so it measures the change in noise at a fixed signal level.
The second column recomputes the same metrics with the true tissue labels,
which you can do only in simulation.

```{r qc}
qc <- compute_design_free_qc(sim$x, fit$x_clean, wNN = fit$wNN,
                             component_table = fit$component_table)
qc_true <- compute_design_free_qc(sim$x, fit$x_clean,
                                  wNN = ifelse(sim$nn, 0, 1))
metrics <- c("tsnr_nt_before", "tsnr_nt_after", "tsnr_nn_before",
             "tsnr_nn_after", "nt_variance_reduction_frac",
             "nn_variance_reduction_frac")
round(cbind(wNN_labels = unlist(qc[metrics]),
            true_labels = unlist(qc_true[metrics])), 3)
```

Median tSNR in non-neuronal voxels more than doubles, and variance in
neuronal voxels falls modestly. The columns disagree for neuronal voxels
because the wNN-labelled neuronal set also contains the
`r sum(sim$nn & fit$wNN > 0.5)` non-neuronal voxels that the map missed, and
denoising removes much of their variance. Their median variance falls by:

```{r qc-missed}
missed <- sim$nn & fit$wNN > 0.5
v_raw <- apply(sim$x[missed, ], 1, var)
v_clean <- apply(fit$x_clean[missed, ], 1, var)
round(1 - median(v_clean) / median(v_raw), 3)
```

These metrics describe what was removed. They can't show that only noise was
removed, because tSNR rises whenever variance is removed, whether it was
noise or neural signal.

`write_qc_artifact()` saves the summary as `.rds`, `.csv` or `.json` for
reports:

```{r qc-write}
path <- tempfile(fileext = ".csv")
write_qc_artifact(qc, path)
head(read.csv(path), 4)
```

## Protecting a task design

For task data you can pass the design matrix. Candidate regressors are then
made orthogonal to the design (plus an intercept) before scoring and removal,
so least-squares task estimates from `x_clean` equal those from the raw data.
To see what that buys and what it costs, add a block-design response
(β = 0.5) to the neuronal voxels only. The design correlates weakly with the
aliased cardiac source:

```{r guardrail-data}
t_sec <- (seq_len(ncol(sim$x)) - 1) * sim$tr
boxcar <- as.numeric(t_sec %% 60 < 30)
hrf <- dgamma(0:15 * sim$tr, shape = 6) - dgamma(0:15 * sim$tr, shape = 16) / 6
task <- as.numeric(stats::filter(boxcar, hrf, sides = 1, circular = TRUE))
task <- task / sd(task)
x_task <- sim$x + outer(ifelse(sim$nn, 0, 0.5), task)
round(cor(task, sim$physio), 3)
```

Fit with and without the design, and estimate the task effect in every voxel
by least squares:

```{r guardrail}
fit_free <- fast_phy_denoise(x_task, tr = sim$tr)
fit_guard <- fast_phy_denoise(x_task, tr = sim$tr, design = task)
X_design <- cbind(1, task)
beta <- function(x) (x %*% X_design %*% solve(crossprod(X_design)))[, 2]
b <- cbind(raw = beta(x_task), no_design = beta(fit_free$x_clean),
           design = beta(fit_guard$x_clean))

# Largest change in any voxel's estimate, relative to the raw data
max_change <- apply(abs(b[, -1] - b[, "raw"]), 2, max)
signif(max_change, 3)

beta_tab <- rbind(
  "mean beta, neuronal voxels (true 0.5)" = colMeans(b[!sim$nn, ]),
  "RMS beta, non-neuronal voxels (true 0)" = sqrt(colMeans(b[sim$nn, ]^2))
)
round(beta_tab, 3)
```

With the design, no estimate changes by more than
`r signif(max_change[["design"]], 2)`, which is numerical precision. Without
it, estimates change by up to `r sprintf("%.2f", max_change[["no_design"]])`.
In non-neuronal voxels, where the true effect is 0, removing the cardiac noise
that correlates with the task shrinks the spurious task effect (RMS
`r round(beta_tab[2, "raw"], 3)` to `r round(beta_tab[2, "no_design"], 3)`).
The guardrail keeps it at `r round(beta_tab[2, "design"], 3)`. The mean
estimate in neuronal voxels shifts slightly without the design
(`r round(beta_tab[1, "raw"], 3)` to `r round(beta_tab[1, "no_design"], 3)`).
So the guardrail protects task estimates from denoising, but it also leaves
task-correlated physiological noise in place. Use it when an unchanged GLM is
the requirement, and omit it when removing task-correlated noise matters
more.

## Key parameters

These are the arguments you're most likely to change:

- `pca_rank` (default 20) caps the rank of the subspace from which components
  are extracted; the rank used is `min(pca_rank, floor(T / 3))`. Larger ranks
  give the lag-1 analysis room to fit spurious components.
- `ratio_thresh` (default 2): a candidate is selected when its
  non-neuronal/neuronal R² ratio exceeds this value. Raise it to remove fewer
  components.
- `pred_thresh` (default `NULL`, off) requires a minimum lag-1 canonical
  correlation. Aliased physiological noise isn't always highly predictable,
  so use it with care (see below).
- `stability = "half"` keeps a candidate only if its spatial R² map is
  reproduced when components are re-extracted in each half of the run
  (`stability_thresh`, default 0.3).
- `extraction_weight` is `"non_neuronal"` by default (weight `1 - wNN`).
  `"neuronal"` gives the PHYCAA+ weighting, and `"none"` gives unweighted data.
  Scoring always uses unweighted data.
- `max_passes` repeats extraction on the cleaned data and removes the union of
  the selected components. One pass is the default.
- `delta_nt`, `delta_nn` and `wnn_threshold` control the tissue map, and
  `wnn` accepts a map you computed yourself (for example from a segmentation).

The TR 2 s dataset above is an easy case: two components are removed, one per
source. The same seed at TR 0.8 s is harder, and there the settings matter:

```{r tuning}
sim08 <- simulate_phy_data(n_vox = 1500, n_time = 300, tr = 0.8, seed = 1)
settings <- list(
  "defaults"           = list(),
  "ratio_thresh = 5"   = list(ratio_thresh = 5),
  "pred_thresh = 0.5"  = list(pred_thresh = 0.5),
  "stability = 'half'" = list(stability = "half"),
  "pca_rank = 10"      = list(pca_rank = 10),
  "pca_rank = 50"      = list(pca_rank = 50),
  "max_passes = 2"     = list(max_passes = 2),
  "PHYCAA+-like"       = phycaa_like
)
tuning <- t(vapply(settings, function(args) {
  f <- do.call(fast_phy_denoise, c(list(sim08$x), args))
  c(removed = ncol(f$regressors), kept(sim08, f$x_clean))
}, numeric(3)))
round(tuning, 3)
```

The default removes `r tuning["defaults", "removed"]` components although
there are only two physiological sources. A closer look at the default fit:

```{r tuning-default}
fit08 <- fast_phy_denoise(sim08$x)
selected08 <- fit08$component_table$selected
# Frequencies the base rates alias to at TR 0.8 s
alias08 <- abs(sim08$physio_hz - round(sim08$physio_hz * sim08$tr) / sim08$tr)
round(alias08, 3)
# Best match between any removed regressor and each true source
round(apply(abs(cor(fit08$regressors, sim08$physio)), 2, max), 2)
# Lag-1 correlation of the selected candidates
round(range(fit08$component_table$predictability[selected08]), 3)
```

At this TR the two sources alias to within
`r round(abs(diff(alias08)), 3)` Hz of each other. No single removed regressor
matches either source closely, yet together the removed components take out
almost all of the physiological energy (first row of the table). Some
selected candidates have almost no lag-1 predictability; they pass on the
spatial ratio alone. Each removed component also takes one temporal degree
of freedom from every voxel, and the three fits that remove the most
components keep the least neural energy.

- `pca_rank = 10` removes fewer components and loses no physiological
  removal.
- `pca_rank = 50` and the PHYCAA+-like settings leave more physiological
  energy behind.
- `pred_thresh = 0.5` removes nothing, because every candidate that passes
  the ratio test has a lag-1 correlation below 0.5.
- A second pass doubles the number of removed components and removes more
  neural signal.

**Masks and input formats.** `mask` restricts the analysis to a set of voxels
(a logical vector, voxel indices, a 3D array, or a mask volume). Voxels
outside it are returned unchanged:

```{r mask}
in_brain <- seq_len(nrow(sim$x)) <= 1200
fit_masked <- fast_phy_denoise(sim$x, mask = in_brain)
identical(fit_masked$x_clean[!in_brain, ], sim$x[!in_brain, ])
```

Voxels that are constant or contain non-finite values are also excluded and
returned unchanged. Input can be a voxels × time matrix, a time × voxels
matrix (set `orientation`, or let `"auto"` treat a matrix with fewer rows than
columns as time × voxels), a 4D array with time last, or a `neuroim2` image.
Matrices and arrays come back with their original dimensions. A
`neuroim2::NeuroVec` comes back as the same kind of image in the same
`NeuroSpace` -- a `DenseNeuroVec`, or a `SparseNeuroVec` with its mask, in
which case only the voxels inside that mask are analysed -- and a
`neuroim2::LogicalNeuroVol` can be used as `mask`:

```{r neuroim2}
sp <- neuroim2::NeuroSpace(c(10, 10, 15, ncol(sim$x)), spacing = c(3, 3, 3))
img <- neuroim2::DenseNeuroVec(array(sim$x, dim(sp)), sp)
fit_img <- fast_phy_denoise(img, tr = sim$tr)
class(fit_img$x_clean)
identical(neuroim2::space(fit_img$x_clean), sp)
```

## Limitations {#limitations}

**No physiological noise, no reliable selection.** The tissue map flags the
voxels with the most excess difference energy. The next chunk runs the same
simulation with the physiological sources switched off, so every truly
non-neuronal voxel is left without physiological noise:

```{r no-physio}
sim0 <- simulate_phy_data(n_vox = 1500, n_time = 300, tr = 2, seed = 1,
                          physio_amp_nn = 0, physio_amp_nt = 0)
fit0 <- fast_phy_denoise(sim0$x)
flagged <- fit0$wNN < 0.5
data.frame(
  removed = ncol(fit0$regressors),
  flagged = sum(flagged),
  flagged_truly_non_neuronal = sum(flagged & sim0$nn),
  neural_kept = round(median(energy_kept(sim0$x[!sim0$nn, ],
                                         fit0$x_clean[!sim0$nn, ],
                                         sim0$neural)), 3)
)
```

The map still flags `r sum(flagged)` voxels as non-neuronal, but only
`r sum(flagged & sim0$nn)` of them is truly non-neuronal. Components expressed
in the flagged neuronal voxels then pass the ratio test:
`r ncol(fit0$regressors)` are removed, and only
`r pct(median(energy_kept(sim0$x[!sim0$nn, ], fit0$x_clean[!sim0$nn, ], sim0$neural)))`
of the neural energy survives. So don't apply the method to data that have
already been physiologically corrected, and inspect `component_table` and the
removed `regressors` (their spectra, their spatial maps) before trusting a
result.

Other limitations:

- **The defaults depart from PHYCAA+.** PHYCAA+ weights toward neuronal
  tissue and searches over subspace ranks. The defaults here (weighting
  toward non-neuronal tissue, rank 20, ratio 2) were chosen because they did
  better on `simulate_phy_data()`, as the replication above shows. That
  evidence comes from simulation, not from validation on acquired data.
- **A physiological source can be only partly removed.** In the worst of the
  six TR 2 s runs above, `r pct(by_source["energy_kept", "cardiac"])` of the
  cardiac energy remained in neuronal voxels.
- **The design guardrail leaves task-correlated noise in place.**
- **CompCor is a baseline.** `compcor_denoise()` follows Behzadi et al.
  (2007) but does not reproduce fMRIPrep's confounds (mask erosion, censoring,
  variance-explained retention). Use fMRIPrep's confounds when you need
  parity.

## Further reading

- `?fast_phy_denoise`: all arguments and the returned object.
- `?compute_wNN_diff_energy` and `?extract_caa_oneshot`: the tissue map and
  the component extractor as standalone functions. The `"dicca"` and
  `"dipca"` extractors (Dong & Qin, 2018) require the 'dipca' package from
  <https://bbuchsbaum.r-universe.dev>.
- `?benchmark_fast_phy_denoise`: timing and ground-truth scoring over data
  sizes and SVD engines.

## References

Behzadi, Y., Restom, K., Liau, J., & Liu, T. T. (2007). A component based
noise correction method (CompCor) for BOLD and perfusion based fMRI.
*NeuroImage*, 37(1), 90–101.
<https://doi.org/10.1016/j.neuroimage.2007.04.042>

Churchill, N. W., & Strother, S. C. (2013). PHYCAA+: An optimized, adaptive
procedure for measuring and controlling physiological noise in BOLD fMRI.
*NeuroImage*, 82, 306–325.
<https://doi.org/10.1016/j.neuroimage.2013.05.102>

Churchill, N. W., Yourganov, G., Spring, R., Rasmussen, P. M., Lee, W.,
Ween, J. E., & Strother, S. C. (2012). PHYCAA: Data-driven measurement and
removal of physiological noise in BOLD fMRI. *NeuroImage*, 59(2), 1299–1314.
<https://doi.org/10.1016/j.neuroimage.2011.08.021>

Dong, Y., & Qin, S. J. (2018). A novel dynamic PCA algorithm for dynamic data
modeling and process monitoring. *Journal of Process Control*, 67, 1–11.
<https://doi.org/10.1016/j.jprocont.2017.05.002>
